Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-19 manuscript/climate_sensitivity_draft.Rmd
Updated with author feedback to end of 2nd paragraph in results
#65

**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest

@@

### Summary
- As climate change is driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important. However, it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts.

- In this study, we analyze tree-ring records for twelve species of an oak-hickory forest -  representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over a 60-year period (1950 - 2009).

- Individual-level drought resistance decreased with tree height, whereas crown position and soil water availability explained little variation. Resistance was lower among species whose leaves had a higher percent loss of area upon dehydration and lost turgor at less negative water potential. However, there was substantial variation in the best predictor variables across the three drought periods. **See comment 1 and 2**

- We conclude that height--as opposed to canopy position, soil water availability, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.

@@

\newpage
### Introduction
**See comment 4**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

@@

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Figs. 1, S2). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3). Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (*Cite Shenandoah paper, if accepted*). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were not identified as a pointer yeasr (Table S3)

>>>>

**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest **see comment 28**

@@

### Summary
- As climate change is driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important. However, it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts. **see comment 29**

- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).

- Individual-level drought resistance decreased with tree height, whereas crown position and soil water availability explained little variation. Resistance was greater both among species whose leaves had a lower turgor loss point, and whose leaves experienced less shrinkage upon dessication. However, there was substantial variation in the best predictor variables across the three drought periods. **See comment 1 and 2**

- We conclude that height--as opposed to canopy position, soil water availability, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts. **see comment 30**

@@

\newpage
### Introduction
**See comment 4,31**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

@@

**Nobby questions using "useful predictor" as a term, instead using more jargon I think**
In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs. 1, S2). We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3). Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp. [*Cite Shenandoah paper, if accepted*; @twery_effects_1991].

Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3). **see comment 32**
